---
output: 
  html_document: 
    fig_caption: yes
    keep_md: yes
---
Reproducible Research: Peer Assessment 2
==========================================

## Impact of Severe Weather Events on Public Health and Economy in the United States

### Synonpsis  
In this report, we aim to analyze the impact of different weather events on public health and economy based on the storm database collected from the U.S. National Oceanic and Atmospheric Administration's (NOAA) from 1950 - 2011. We will use the estimates of fatalities, injuries, property and crop damage to decide which types of event are most harmful to the population health and economy. From these data, we found that excessive heat and tornado are most harmful with respect to population health, while flood, drought, and hurricane/typhoon have the greatest economic consequences.

### Basic settings
```{r}
echo = TRUE  # Always make code visible
options(scipen = 1)  # Turn off scientific notations for numbers

library(ggplot2)
library(plyr)

suppressWarnings(suppressMessages(library(R.utils)))
suppressWarnings(suppressMessages(require(gridExtra)))
```

### Data Processing
First, we download the data file and unzip it.
```{r}

if (!"StormData.csv.bz2" %in% dir("~/RData/")) {
    print("hhhh")
    download.file("http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2", destfile = "RData/stormData.csv.bz2")
    bunzip2("RData/stormData.csv.bz2", overwrite=T, remove=F)
}

```
Then, we read the generated csv file. If the data already exists in the working environment, we do not need to load it again. Otherwise, we read the csv file.
```{r}
if (!"stormData" %in% ls()) {
    stormData <- read.csv("RData/stormData.csv", sep = ",")
}
dim(stormData)
head(stormData, n = 2)
```
There are 902297 rows and 37 columns in total.
The events in the database start in the year 1950 and end in November 2011. In the earlier years of the database there are generally fewer events recorded, most likely due to a lack of good records. More recent years should be considered more complete.

```{r}
if (dim(stormData)[2] == 37) {
    stormData$year <- as.numeric(format(as.Date(stormData$BGN_DATE, format = "%m/%d/%Y %H:%M:%S"), "%Y"))
}
hist(stormData$year, breaks = 30, main = "Histogram of Storm Data")
```  

Based on the above histogram, we see that the number of events tracked starts to significantly increase around 1995. So, we use the subset of the data from 1990 to 2011 to get most out of good records.
```{r}
storm <- stormData[stormData$year >= 1995, ]
dim(storm)
```
Now, there are 681500 rows and 38 columns in total.

#### Impact on Public Health
In this section, we check the number of **fatalities** and **injuries** that are caused by the severe weather events. We would like to get the first 15 most severe types of weather events.
```{r}
sortHelper <- function(fieldName, top = 15, dataset = stormData) {
    index <- which(colnames(dataset) == fieldName)
    field <- aggregate(dataset[, index], by = list(dataset$EVTYPE), FUN = "sum")
    names(field) <- c("EVTYPE", fieldName)
    field <- arrange(field, field[, 2], decreasing = T)
    field <- head(field, n = top)
    field <- within(field, EVTYPE <- factor(x = EVTYPE, levels = field$EVTYPE))
    return(field)
}

```

#### Impact on Economy
We will convert the **property damage** and **crop damage** data into comparable numerical forms according to the meaning of units described in the code book ([Storm Events](http://ire.org/nicar/database-library/databases/storm-events/)). Both `PROPDMGEXP` and `CROPDMGEXP` columns record a multiplier for each observation where we have Hundred (H), Thousand (K), Million (M) and Billion (B).

```{r}
convertHelper <- function(dataset = storm, fieldName, newFieldName) {
    totalLen <- dim(dataset)[2]
    index <- which(colnames(dataset) == fieldName)
    dataset[, index] <- as.character(dataset[, index])
    logic <- !is.na(toupper(dataset[, index]))
    dataset[logic & toupper(dataset[, index]) == "B", index] <- "9"
    dataset[logic & toupper(dataset[, index]) == "M", index] <- "6"
    dataset[logic & toupper(dataset[, index]) == "K", index] <- "3"
    dataset[logic & toupper(dataset[, index]) == "H", index] <- "2"
    dataset[logic & toupper(dataset[, index]) == "", index] <- "0"
    dataset[, index] <- as.numeric(dataset[, index])
    dataset[is.na(dataset[, index]), index] <- 0
    dataset <- cbind(dataset, dataset[, index - 1] * 10^dataset[, index])
    names(dataset)[totalLen + 1] <- newFieldName
    return(dataset)
}

storm <- convertHelper(storm, "PROPDMGEXP", "propertyDamage")
storm <- convertHelper(storm, "CROPDMGEXP", "cropDamage")
names(storm)
options(scipen=999)
```

Now I'm going to clean up some of the obvious event_types errors.

```{r textreplacement}
storm$EVTYPE <- toupper(storm$EVTYPE)
# remove leading and traling spaces
storm$EVTYPE <- gsub("^[[:space:]]+|[[:space:]]+$", "", storm$EVTYPE)
# make more uniform the used vocabulary
storm$EVTYPE <- gsub("EXCESSIVE|EXCESSIVELY|EXTREMELY", "EXTREME", storm$EVTYPE)
storm$EVTYPE <- gsub("FLOODINGINGS*|FLOOODING|FLOOD", "FLOODING", storm$EVTYPE)
storm$EVTYPE <- gsub("FLASH FLOODING.*", "FLOODING", storm$EVTYPE)
storm$EVTYPE <- gsub("LIGHTNING\\.|LIGHTING|LIGNTNING", "LIGHTNING", storm$EVTYPE)
storm$EVTYPE <- gsub("NON-", "NON ", storm$EVTYPE)
storm$EVTYPE <- gsub("PROLONG", "PROLONGED", storm$EVTYPE)
storm$EVTYPE <- gsub("RAINS", "RAIN", storm$EVTYPE)
storm$EVTYPE <- gsub("RIP CURRENTS", "RIP CURRENT", storm$EVTYPE)
storm$EVTYPE <- gsub("STORMS", "STORM", storm$EVTYPE)
storm$EVTYPE <- gsub("TORNDAO|TORNADOES", "TORNADO", storm$EVTYPE)
storm$EVTYPE <- gsub("TH*UND*ER*[A-Z]*RMW*|THUNDERSTROM|THUDERSTORM",
                    "THUNDERSTORM", storm$EVTYPE)
storm$EVTYPE <- gsub("TSTM WIND","THUNDERSTORM WIND", storm$EVTYPE)
storm$EVTYPE <- gsub("UNUSUALLY", "UNUSUAL", storm$EVTYPE)
storm$EVTYPE <- gsub("WILD.*FIRE.*|WILD/FOREST.*", "WILD/FOREST FIRES", storm$EVTYPE)
storm$EVTYPE <- gsub("WINDS|WND", "WIND", storm$EVTYPE)
storm$EVTYPE <- gsub("WINTERY", "WINTER", storm$EVTYPE)
storm$EVTYPE <- gsub("WARMTH", "WARM", storm$EVTYPE)
# grouping some events
storm$EVTYPE <- gsub("^BLIZZARD.*|ICE STORM", "BLIZZARD", storm$EVTYPE)
storm$EVTYPE <- gsub("^COASTAL.*|.*/CSTL .*", "COASTAL EROSION/FLOODING/STORM ",
                    storm$EVTYPE)
storm$EVTYPE <- gsub("EXTREME COLD.*|EXTENDED COLD.*", "EXTREME COLD", storm$EVTYPE)
storm$EVTYPE <- gsub("^DRY.*", "DRY CONDITIONS", storm$EVTYPE)
storm$EVTYPE <- gsub("^FLOODING.*", "FLOODING", storm$EVTYPE)
storm$EVTYPE <- gsub("^FREEZE|^FREEZING.*|^FROST.*",
                    "FREEZING FOG/RAIN/SLEET/SNOW", storm$EVTYPE)
storm$EVTYPE <- gsub("HAIL.*", "HAIL", storm$EVTYPE)
storm$EVTYPE <- gsub("DROUGHT|EXTREME HEAT.*|^HEAT.*", "EXTREME HEAT", storm$EVTYPE)
storm$EVTYPE <- gsub("HEAVY RAIN.*", "HEAVY RAIN", storm$EVTYPE)
storm$EVTYPE <- gsub("HURRICANE.*", "HURRICANE", storm$EVTYPE)
storm$EVTYPE <- gsub("HEAVY SNOW.*|^SNOW.*|EXCESSIVE SNOW", "HEAVY SNOW/ICE",
                    storm$EVTYPE)
storm$EVTYPE <- gsub("LIGHTNING.*", "LIGHTNING", storm$EVTYPE)
storm$EVTYPE <- gsub("^MARINE.*", "MARINE THUNDERSTORM/ACCIDENT", storm$EVTYPE)
storm$EVTYPE <- gsub("RAIN.*|PROLONGEDED RAIN", "RAIN", storm$EVTYPE)
storm$EVTYPE <- gsub("RIP CURRENT.*|HEAVY SURF.*|HIGH SURF.*", "HEAVY SURF", storm$EVTYPE)
storm$EVTYPE <- gsub("SLEET.*", "SLEET", storm$EVTYPE)
storm$EVTYPE <- gsub("VOLCANIC.*", "VOLCANIC", storm$EVTYPE)
storm$EVTYPE <- gsub("THUNDERSTORM.*|SEVERE THUNDERSTORM", "THUNDERSTORM", storm$EVTYPE)
storm$EVTYPE <- gsub("TORNADO.*", "TORNADO", storm$EVTYPE)
storm$EVTYPE <- gsub("TROPICAL STORM.*", "TROPICAL STORM", storm$EVTYPE)
storm$EVTYPE <- gsub("UNSEASONAL.*|^UNSEASONABL[EY].*|^^UNUSUAL.*",
                    "UNUSUAL WEATHER CONDITION", storm$EVTYPE)
storm$EVTYPE <- gsub("HIGH WIND.*|STRONG WIND.*|^WIND.*", "HIGH WIND", storm$EVTYPE)
storm$EVTYPE <- gsub("^WATERSPOUT.*|WATER SPOUT", "WATERSPOUT", storm$EVTYPE)
storm$EVTYPE <- gsub("^WINTER.*", "WINTER STORM/WIND", storm$EVTYPE)
storm$EVTYPE <- gsub("^NONE|^SUMMARY.*", "?", storm$EVTYPE)

# remove multiple spaces
storm$EVTYPE <- gsub("[[:space:]]+", " ", storm$EVTYPE)

```


### Results
As for the impact on public health, we have got two sorted lists of severe weather events below by the number of people badly affected.
```{r}
fatalities <- sortHelper("FATALITIES", dataset = storm)
injuries <- sortHelper("INJURIES", dataset = storm)

fatalities
injuries
```
And the following is a pair of graphs of total fatalities and total injuries affected by these severe weather events. 
```{r}
fatalitiesPlot <- qplot(EVTYPE, data = fatalities, weight = FATALITIES, geom = "bar") + 
    scale_y_continuous("Number of Fatalities") + 
    theme(axis.text.x = element_text(angle = 45, 
    hjust = 1)) + xlab("Severe Weather Type") + 
    ggtitle("Total Fatalities by Severe Weather\n Events in the U.S.\n from 1995 - 2011")
injuriesPlot <- qplot(EVTYPE, data = injuries, weight = INJURIES, geom = "bar") + 
    scale_y_continuous("Number of Injuries") + 
    theme(axis.text.x = element_text(angle = 45, 
    hjust = 1)) + xlab("Severe Weather Type") + 
    ggtitle("Total Injuries by Severe Weather\n Events in the U.S.\n from 1995 - 2011")
grid.arrange(fatalitiesPlot, injuriesPlot, ncol = 2)
```  

Based on the above histograms, we find that **Extreme Heat** and **Tornado** cause most fatalities and injuries in the United States from 1995 to 2011.

As for the impact on economy, we have got two sorted lists below by the amount of money cost by damages.  

```{r}
property <- sortHelper("propertyDamage", dataset = storm)
crop <- sortHelper("cropDamage", dataset = storm)

property
crop
```
And the following is a pair of graphs of total property damage and total crop damage affected by these severe weather events. 
```{r}
propertyPlot <- qplot(EVTYPE, data = property, weight = propertyDamage, geom = "bar") + 
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) + scale_y_continuous("Property Damage in US dollars")+ 
    xlab("Severe Weather Type") + ggtitle("Total Property Damage by\n Severe Weather Events in\n the U.S. from 1995 - 2011")

cropPlot<- qplot(EVTYPE, data = crop, weight = cropDamage, geom = "bar") + 
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) + scale_y_continuous("Crop Damage in US dollars") + 
    xlab("Severe Weather Type") + ggtitle("Total Crop Damage by \nSevere Weather Events in\n the U.S. from 1995 - 2011")
grid.arrange(propertyPlot, cropPlot, ncol = 2)
```  

Based on the above histograms, we find that **Flooding** and **Hurricane** cause most property damage; **Extreme Heat** and **Flooding** causes most crop damage in the United States from 1995 to 2011.

### Conclusion  
From this data we can see that **Hurricane** and **Extreme Heat** causes the most damage to all sources with **Flooding** coming in second place. 